---
date: '`r Sys.Date()`'
output:
  html_document:
    toc: true
params:
  invnorm: FALSE
  n_categories_boxplot: 10
  pipeline_version: 0
---

---
title: "Null model report `r if (!is.null(params$invnorm) & params$invnorm) "- rank-based inverse normal transform"`"
author: "TOPMed Pipeline v`r params$pipeline_version`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, results = "hide")
library(TopmedPipeline)
library(GENESIS)
library(Biobase)
library(dplyr)
library(ggplot2)
library(grid)
library(stringr)
theme_set(theme_bw())

configTable <- function(x) {
    x <- x[!is.na(x)]
    dat <- data.frame(parameter=names(x), value=x, row.names=1:length(x))
    knitr::kable(dat)
}

COLOR_ZERO_LINE = "grey60"
n_categories_boxplot = as.numeric(params$n_categories_boxplot)
```

```{r}
files <- list.files(pattern=".+params$")
scripts <- sapply(files, function(x) {
    tmp <- strsplit(x, split=".", fixed=TRUE)[[1]]
    tmp[length(tmp)-1]
})
param_files <- lapply(setNames(files, scripts), readConfig)
```

```{r}
config <- param_files$null_model
phen <- getPhenotypes(config)
outcome <- phen$outcome
covars <- phen$covars
annot <- filter(pData(phen$annot), sample.id %in% phen$sample.id)
random <- if (!is.na(config["relatedness_matrix_file"])) "relatedness" else NULL
group_var <- if (is.na(config["group_var"])) NULL else config["group_var"]
nullmod_file <- sprintf("%s_reportonly.RData", config["out_prefix"])
if (params$invnorm) {
  nullmod_file <- sprintf("%s_invnorm_reportonly.RData", config["out_prefix"])
}
nullmod <- getobj(nullmod_file)
is_binary <- config["family"] == "binomial"

is_invnorm <- !is_binary & params$invnorm & as.logical(config["inverse_normal"])

model_string <- nullmod$model$formula
outcome_string <- str_trim(str_split(model_string, "~")[[1]][1])
```

```{r prepare-data}
# Display any covariate with < 10 values using boxplots instead of scatterplots.
for (covar in covars) {
  if (!is.numeric(annot[[covar]]) || length(unique(annot[[covar]])) < n_categories_boxplot) {
    annot[[covar]] <- as.factor(annot[[covar]])
  }
}

# Set phenotype file outcome to NULL to ensure we use the null model object.
annot[[outcome]] <- NULL

dat <- nullmod$fit
# Rename for clearer plotting.
names(dat)[names(dat) == "outcome"] <- "model_outcome"
dat$outcome_numeric <- dat$model_outcome
if (is_binary) {
  # Convert to factor for plotting binary outcomes.
  dat$model_outcome <- as.factor(dat$model_outcome)
}

# TODO: Add group_var check.
g <- length(nullmod$group.idx)
group <- rep(NA, length(nullmod$fit$sample.id))
for(i in seq_len(g)){
    group[nullmod$group.idx[[i]]] <- names(nullmod$group.idx)[i]
}
dat$group <- group
dat <- left_join(dat, annot, by = "sample.id")

```

## Config

Number of samples: `r nrow(annot)`

`r model_string`

```{r, results='asis'}
disp <- c("sample_include_file", "phenotype_file", "pca_file", "relatedness_matrix_file")
if (as.logical(config["inverse_normal"])) {
    disp <- c(disp, "resid_covars")
    if (!is.na(config["group_var"]) & params$invnorm) {
        disp <- c(disp, c("inverse_normal", "rescale_variance"))
    }
}
configTable(config[disp])
```

## Phenotypes

### Distributions

```{r outcome-distribution-plot}
if (is_binary) {
    p <- ggplot(dat, aes(model_outcome)) +
      geom_bar() +
      stat_count(geom="text", aes_string(label="..count..", vjust=-0.5)) +
      xlab(outcome_string)
} else {
    p <- ggplot(dat, aes(model_outcome)) +
      geom_histogram() +
      xlab(outcome_string)
}
print(p)
```

```{r covariate-distribution-plot}
for (x in covars[-grep("^PC[[:digit:]]+", covars)]) {
    if (is.numeric(annot[[x]])) {
        print(ggplot(annot, aes_string(x)) + geom_histogram())
    } else {
        print(ggplot(annot, aes_string(x)) + geom_bar() + stat_count(geom="text", aes_string(label="..count..", vjust=-0.5)))
    }
}
```

## Covariate effect size estimates

Covariates are considered to be significant if they have $p < 0.05$ after a Bonferroni correction for the number of fixed effects.

```{r fixef-plot}
fixef <- nullmod$fixef
# Rename fixef columns because they are different for different types of models.
names(fixef) <- c("Est", "SE", "Stat", "pval")
fixef$covar <- rownames(fixef)
fixef <- fixef %>%
  mutate(
    significant = pval < (0.05 / n()),
    errorbar_min = Est + qnorm(0.025)*SE,
    errorbar_max = Est + qnorm(0.975)*SE
  )
cols <- c("FALSE" = "grey70", "TRUE" = "red")
p <- ggplot(fixef, aes(x = Est, y = covar, color = significant)) +
    geom_vline(xintercept = 0, color = COLOR_ZERO_LINE) +
    geom_point() +
    geom_errorbarh(aes(xmin = errorbar_min, xmax = errorbar_max)) +
    scale_color_manual(values = cols)

print(p)
```


```{r fixef-plot-signif}
if (any(fixef$significant)) {
  # Make a plot that's zoomed in to show only the significant effects.
  idx_signif <- fixef[, "significant"]
  # Calculate the range based on the error bars.
  minx <- min(c(0, fixef$errorbar_min[idx_signif]))
  maxx <- max(c(0, fixef$errorbar_max[idx_signif]))
  p2 <- p +
     coord_cartesian(xlim = c(minx, maxx)) +
     ggtitle("Covariate effects - zoomed to significant only")
  print(p2)
}
```

```{r fixef-table-significant, results = 'asis'}
if (any(fixef$significant)) {
  cat('### Significant covariate effect sizes\n\n')
  table_data <- fixef[fixef$significant, c("covar", "Est", "SE", "Stat", "pval")]
  table_data$pval <- sprintf("%4.3e", table_data$pval)
  names(table_data)[[1]] <- "Covariate"
  k <- knitr::kable(table_data, align = "r")
  print(k)
}
```

## Marginal residuals


```{r marginal-helper-text, results = "asis"}
if (is_binary) {
  text <- paste(
    "For continuous covariates, plots of marginal residuals vs covariates
    should have a red linear trend line near the `y = 0` line and a blue
    smoothed (cubic regression spline) curve near the y = 0 line. For
    categorical covariates, the mean of each group should be near 0. Linear
    covariates with a small number of unique values are shown using boxplots.
    A distinct trend could indicate that a more complex term for that covariate
    is needed (e.g. quadratic, spline, etc.)."
  )
} else {
  text <- paste(
    "Plots of marginal residuals vs covariates should show no distinct
    pattern, appearing as random noise. For continuous covariates, the red
    linear trend line and the blue smoothed curve (cubic regression spline)
    should be near the y = 0 line. For categorical covariates, the black cross
    indicating the group mean should be near 0. Linear covariates with a small
    number of unique values are shown using boxplots. A distinct pattern or
    trend could indicate that a more complex term for that covariate is needed,
    (e.g. quadratic, spline, etc.)."
  )
}
text <- paste(text, "Plots of the outcome versus each covariate are also shown for reference.")
cat(text)
```

```{r residual-plots, results = "asis"}
for (x in covars) {
    cat(sprintf("\n\n### %s {.unlisted .unnumbered}\n\n", x))
    p_out <- ggplot(dat, aes_string(x = x, y = "outcome_numeric")) +
      ggtitle(outcome_string) +
      ylab(outcome_string)
    p_res <- ggplot(dat, aes_string(x = x, y = "resid.marginal")) +
      ggtitle(sprintf("%s - residuals", outcome_string))
    if (is.numeric(annot[[x]] )) {
      p_res <- p_res +
        geom_point(alpha = 0.1) +
        geom_hline(yintercept = 0, color = COLOR_ZERO_LINE) +
        geom_smooth(method = 'lm', color = 'red', fill = 'red') +
        geom_smooth(color = 'blue', linetype = 'dashed', fill = 'blue')
      p_out <- p_out +
        geom_point(alpha = 0.1) +
        #geom_hline(yintercept = 0, color = COLOR_ZERO_LINE) +
        geom_smooth(method = 'lm', color = 'red', fill = 'red') +
        geom_smooth(color = 'blue', linetype = 'dashed', fill = 'blue')
    } else {
      p_res <- p_res +
        geom_boxplot(varwidth = TRUE) +
        geom_hline(yintercept = 0, color = COLOR_ZERO_LINE) +
        # Means by group.
        stat_summary(fun.y=mean, color="black", shape = 4, geom="point", size = 3) +
        # Rotate axis labels.
        theme(axis.text.x = element_text(angle = 30, vjust = 1, hjust = 1))
      p_out <- p_out +
        geom_boxplot(varwidth = TRUE) +
        #geom_hline(yintercept = 0, color = COLOR_ZERO_LINE) +
        ## Means by group.
        stat_summary(fun.y=mean, color="black", shape = 4, geom="point", size = 3) +
        # Rotate axis labels.
        theme(axis.text.x = element_text(angle = 30, vjust = 1, hjust = 1))
    }
    print(p_out)
    print(p_res)
}
```

```{r, results = 'asis'}
if (is_binary) {
  text <- paste(
    "Plots of marginal residuals vs fitted values should have the red linear
    trend line and the blue smoothed curve (cubic regression spline) near the
    `y = 0` line. A distinct pattern or trend could indicate that additional
    covariates are needed, or that the variance model is inadequate."
  )
} else {
  text <- paste(
    "Plots of marginal residuals vs fitted values should show no distinct
    pattern, appearing as random noise, with the red linear trend line and the
    blue smoothed curve (cubic regression spline) near the `y = 0` line. A
    distinct pattern or trend could indicate that additional covariates are
    needed, or that the variance model is inadequate."
  )
}
cat(text)
```

```{r}
p <- ggplot(dat, aes(x = fitted.values, y = resid.marginal)) +
    geom_point(alpha = 0.1) +
    geom_hline(yintercept = 0, color = COLOR_ZERO_LINE) +
    geom_smooth(method = 'lm', color = 'red', fill = 'red') +
    geom_smooth(color = 'blue', linetype = 'dashed', fill = 'blue')

print(p)
```

```{r varcomp-header, results = 'asis'}
show_varcomp <- FALSE
if (!is.null(random)){
  cat("## Variance components\n")

  if (!all(nullmod$zeroFLAG)) {
    show_varcomp <- TRUE
  } else {
    cat("\nAll variance components are estimated to be 0.\n\n")
  }
}
```

```{r varcomp}
if (show_varcomp) {
  vc <- varCompCI(nullmod, prop = FALSE)
  vc$component <- rownames(vc)
    p <- ggplot(vc, aes(x = Est, y = component)) +
      geom_vline(xintercept = 0, color = COLOR_ZERO_LINE) +
      geom_point() +
      geom_errorbarh(aes(xmin = `Lower 95`, xmax = `Upper 95`))
    print(p)
}
```

```{r varcomp-table, results = 'asis'}
if (show_varcomp) {
  cat("### Variance component estimates\n\n")
  k <- knitr::kable(vc)
  print(k)
}
```

## Adjusted phenotype

The Cholesky residuals (`resid.cholesky`) are adjusted for fixed effect covariates,
random effects, and heterogeneous residual variances (if applicable).
```{r results = 'asis'}
if (is_binary) {
  cat("Linear regression of the Cholesky residual vector on an equivalently
       transformed genotype vector provides the same estimates as fitting the
       full GLS model using `workingY` as the outcome variable (by the
        Frisch-Waugh-Lovell theorem). The Cholesky residuals can be viewed as
        the adjusted phenotype values.")
} else {
  cat("Linear regression of the Cholesky residual vector on an equivalently
       transformed genotype vector provides the same estimates as fitting the
       full GLS model (by the Frisch-Waugh-Lovell theorem). The Cholesky
       residuals can be viewed as the adjusted phenotype values.")
}
```

The Cholesky residuals are a transformation of the marginal residuals using the
estimated model covariance structure such that they should be uncorrelated.
```{r, results = 'asis'}
extra_text <- ""
if (is_binary) {
  extra_text <- "A histogram color-coded by outcome status is shown below."
} else {
  extra_text <- "For linear outcomes, they follow a standard multivariate
                 Normal distribution with mean 0 and identity covariance matrix
                asymptotically."
  if (!is.null(group_var)) {
    extra_text <- paste(extra_text, "The density curve for each group is shown; the heavy black
                   line is the density curve pooled across all groups.")
  } else {
    extra_text <- paste(extra_text, "A histogram of `resid.cholesky` values is shown below.")
  }
  if(!is_invnorm) {
    extra_text <- paste(
      extra_text,
      "\n\nIt may be problematic for association testing if this distribution is far
      from Gaussian. This problem can be alleviated by rank normalizing `resid.marginal` and
      re-fitting the null model with the rank-normalized values as the outcome
      (i.e. a two stage procedure)."
    )
  }
}
cat(extra_text)
```

```{r, fig.height=8}
p <- ggplot(dat, aes(x = resid.cholesky))
if (!is.null(group_var)) {
  p <- p +
    stat_density(aes(color = group), geom = "line", position = "identity") +
    #geom_density(aes(color = group)) +
    geom_density(size = 2) +
    theme(legend.position = "top", legend.justification = c(0.5, 1)) +
    guides(color = guide_legend(title.position="top", title.hjust = 0.5))
  print(p)
} else {
  if (is_binary) {
    p <- p + geom_histogram(aes_string(fill = "model_outcome"))
  } else {
    p <- p + geom_histogram()
  }
  print(p)
}
```

```{r variance-adjustment-header, results = 'asis'}
if (!is.null(group_var)) {
  cat(
    "Boxplots of the original phenotype values (`workingY`) and the Cholestky
    residuals (`resid.cholesky`) by heterogeneous residual variance group. The
    variation of `resid.cholesky` within each group should be similar."
  )

  p <- ggplot(dat, aes(x = group, y = workingY)) +
    geom_boxplot(varwidth = TRUE) +
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    # Means by group.
    stat_summary(fun.y=mean, color="black", shape = 4, size = 3, geom="point") +
    # Rotate axis labels.
    theme(axis.text.x = element_text(angle = 30, vjust = 1, hjust = 1))
  print(p)

  p <- ggplot(dat, aes(x = group, y = resid.cholesky))
  p <- p +
    geom_boxplot(varwidth = TRUE) +
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    geom_hline(yintercept = 0, color = COLOR_ZERO_LINE) +
    stat_summary(fun.y=mean, color="black", shape = 4, size = 3, geom="point") +
    # Rotate axis labels.
    theme(axis.text.x = element_text(angle = 30, vjust = 1, hjust = 1))
  print(p)

}
```


The square of these values should have a trend line near the y = 1 line, indicating homoskedasticity after adjustment.
Deviation from this trend could indicate that the variance model is inadequate.

```{r adjusted-phenotype-plots}

p <- ggplot(dat, aes(x = fitted.values, y = resid.cholesky^2)) +
  geom_point(alpha = 0.1) +
  geom_hline(yintercept = 1, color = COLOR_ZERO_LINE) +
  geom_smooth(method = 'lm', color = 'red', fill = 'red') +
  geom_smooth(color = 'blue', linetype = 'dashed', fill = 'blue')
print(p)

p <- p +
  coord_cartesian(ylim = c(0, 5)) +
  ylab("resid.cholesky^2 zoomed in")
print(p)
```

# Session information

```{r sessionInfo(), results = 'markup'}
library(GENESIS)
print(sessionInfo())
```
