---
date: '`r Sys.Date()`'
output:
  html_document:
    toc: true
params:
  invnorm: FALSE
  pipeline_version: 0
---

---
title: "Null model report `r if (!is.null(params$invnorm) & params$invnorm) "- rank-based inverse normal transform"`"
author: "TOPMed Pipeline v`r params$pipeline_version`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, results = "hide")
library(TopmedPipeline)
library(GENESIS)
library(Biobase)
library(dplyr)
library(ggplot2)
theme_set(theme_bw())

configTable <- function(x) {
    x <- x[!is.na(x)]
    dat <- data.frame(parameter=names(x), value=x, row.names=1:length(x))
    knitr::kable(dat)
}

COLOR_ZERO_LINE = "grey60"
```

```{r}
files <- list.files(pattern=".+params$")
scripts <- sapply(files, function(x) {
    tmp <- strsplit(x, split=".", fixed=TRUE)[[1]]
    tmp[length(tmp)-1]
})
param_files <- lapply(setNames(files, scripts), readConfig)
```

```{r}
config <- param_files$null_model
phen <- getPhenotypes(config)
outcome <- phen$outcome
covars <- phen$covars
annot <- filter(pData(phen$annot), sample.id %in% phen$sample.id)
random <- if (!is.na(config["pcrelate_file"])) "kinship" else if (!is.na(config["grm_file"])) "grm" else NULL
group_var <- if (is.na(config["group_var"])) NULL else config["group_var"]
model.string <- modelString(outcome, covars, random, group_var)
nullmod_file <- gsub(".RData", "_small.RData", config["out_file"])
if (params$invnorm) {
  nullmod_file <- gsub(".RData", "_invnorm_small.RData", config["out_file"])
}
nullmod <- getobj(nullmod_file)
is_binary <- as.logical(config["binary"])
```

## Config

Number of samples: `r nrow(annot)`

`r model.string`

```{r, results='asis'}
disp <- c("sample_include_file", "phenotype_file", "pca_file", "pcrelate_file", "grm_file")
if (as.logical(config["inverse_normal"])) {
    disp <- c(disp, "resid_covars")
    if (!is.na(config["group_var"])) {
        disp <- c(disp, "rescale_variance")
    }
}
configTable(config[disp])
```

## Phenotype distributions

```{r outcome-distribution-plot}
outcome_dat <- data.frame(x = nullmod$outcome)
names(outcome_dat)[1] <- outcome

if (is_binary) {
    ggplot(outcome_dat, aes_string(outcome)) + geom_bar() + stat_count(geom="text", aes_string(label="..count..", vjust=-0.5))
} else {
    ggplot(outcome_dat, aes_string(outcome)) + geom_histogram()
}
```

```{r covariate-distribution-plot}
for (x in covars[-grep("^PC[[:digit:]]+", covars)]) {
    if (is.numeric(annot[[x]] )) {
        print(ggplot(annot, aes_string(x)) + geom_histogram())
    } else {
        print(ggplot(annot, aes_string(x)) + geom_bar() + stat_count(geom="text", aes_string(label="..count..", vjust=-0.5)))
    }
}
```

## Covariate effect size estimates

```{r fixef-plot}
fixef <- nullmod$fixef
# Rename fixef columns because they are different for different types of models.
names(fixef) <- c("Est", "SE", "Stat", "pval")
fixef$covar <- rownames(fixef)
fixef <- fixef %>%
  mutate(
    significant = pval < (0.05 / n()),
    errorbar_min = Est + qnorm(0.025)*SE,
    errorbar_max = Est + qnorm(0.975)*SE
  )
cols <- c("FALSE" = "grey70", "TRUE" = "red")
p <- ggplot(fixef, aes(x = Est, y = covar, color = significant)) +
    geom_vline(xintercept = 0, color = COLOR_ZERO_LINE) +
    geom_point() +
    geom_errorbarh(aes(xmin = errorbar_min, xmax = errorbar_max)) +
    scale_color_manual(values = cols)

print(p)
```


```{r fixef-plot-signif}
if (any(fixef$significant)) {
  # Make a plot that's zoomed in to show only the significant effects.
  idx_signif <- fixef[, "significant"]
  # Calculate the range based on the error bars.
  minx <- min(c(0, fixef$errorbar_min[idx_signif]))
  maxx <- max(c(0, fixef$errorbar_max[idx_signif]))
  p2 <- p +
     coord_cartesian(xlim = c(minx, maxx)) +
     ggtitle("Covariate effects - zoomed to significant only")
  print(p2)
}
```

## Marginal residuals


```{r marginal-helper-text, results = "asis"}
if (is_binary) {
  text <- paste(
    "Plots of marginal residuals vs covariates should have a trend line near
    the `y = 0` line for continuous covariates, or the mean of each group near
    0 for categorical covariates. A distinct trend could indicate that a more
    complex term for that covariate is needed (e.g. quadratic, spline, etc.)."
  )
} else {
  text <- paste(
    "Plots of marginal residuals vs covariates should show no distinct
    pattern, appearing as random noise, with the trend line near the `y = 0`
    line for continuous covariates, or the mean of each group near 0 for
    categorical covariates. A distinct pattern or trend could indicate that a
    more complex term for that covariate is needed, (e.g. quadratic, spline,
    etc.)."
  )
}
cat(text)
```

```{r residual-plots}
dat <- data.frame(sample.id = nullmod$sample.id,
                  workingY = nullmod$workingY,
                  fitted.values = nullmod$fitted.values,
                  resid.marginal = nullmod$resid.marginal,
                  PY = as.vector(nullmod$resid),
                  Ytilde = as.vector(nullmod$Ytilde),
                  stringsAsFactors = FALSE)
# TODO: Add group_var check.
g <- length(nullmod$group.idx)
group <- rep(NA, length(nullmod$sample.id))
for(i in 1:g){
    group[nullmod$group.idx[[i]]] <- names(nullmod$group.idx)[i]
}
dat$group <- group
dat <- left_join(dat, annot, by = "sample.id")

for (x in covars) {
    p <- ggplot(dat, aes_string(x = x, y = "resid.marginal"))
    if (is.numeric(annot[[x]] )) {
      p <- p +
        geom_point(alpha = 0.1) +
        geom_hline(yintercept = 0, color = COLOR_ZERO_LINE) +
        geom_smooth(method = 'lm', color = 'red', fill = 'red') +
        geom_smooth(color = 'blue', linetype = 'dashed', fill = 'blue')
    } else {
      p <- p +
        geom_boxplot() +
        geom_hline(yintercept = 0, color = COLOR_ZERO_LINE) +
        # Means by group.
        stat_summary(fun.y=mean, color="red", geom="point")
    }
    print(p)
}
```

```{r, results = 'asis'}
if (is_binary) {
  text <- paste(
    "Plots of marginal residuals vs fitted values should have a trend line near
    the `y = 0` line. A distinct pattern or trend could indicate that additional
    covariates are needed, or that the variance model is inadequate."
  )
} else {
  text <- paste(
    "Plots of marginal residuals vs fitted values should show no distinct
    pattern, appearing as random noise, with the trend line near the `y = 0`
    line. A distinct pattern or trend could indicate that additional covariates
    are needed, or that the variance model is inadequate."
  )
}
cat(text)
```

```{r}
p <- ggplot(dat, aes(x = fitted.values, y = resid.marginal)) +
    geom_point(alpha = 0.1) +
    geom_hline(yintercept = 0, color = COLOR_ZERO_LINE) +
    geom_smooth(method = 'lm', color = 'red', fill = 'red') +
    geom_smooth(color = 'blue', linetype = 'dashed', fill = 'blue')

print(p)
```

```{r varcomp-header, results = 'asis'}
if (!is.null(random)){
  cat("## Variance components")
}
```

```{r varcomp}
if (!is.null(random)) {
  vc <- varCompCI(nullmod, prop = FALSE)
  vc$component <- rownames(vc)
  p <- ggplot(vc, aes(x = Est, y = component)) +
    geom_vline(xintercept = 0, color = COLOR_ZERO_LINE) +
    geom_point() +
    geom_errorbarh(aes(xmin = `Lower 95`, xmax = `Upper 95`))
  print(p)
}
```

```{r variance-adjustment-header, results = 'asis'}
if (!is.null(group_var)) {
  cat("## Variance adjustment")
}
cat('\n\n')
text <- paste(
  "Boxplots of the original phenotype values (`Y`) and the adjusted phenotype
  values (`Ytilde`) by heterogeneous residual variance group. The `Ytilde` values
  are adjusted for fixed effect covariates, random effects, and heterogeneous
  residual variances. The variation of `Ytilde` within each group should be
  similar."
)
cat(text)
```

```{r variance-adjustment-plots}
if (!is.null(group_var)) {

  p <- ggplot(dat, aes(x = group, y = workingY)) +
    geom_boxplot() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    geom_hline(yintercept = 0, color = COLOR_ZERO_LINE) +
    # Means by group.
    stat_summary(fun.y=mean, color="red", geom="point") +
    ylab("Y")
  print(p)

  if (!is.null(group_var)) {
    p <- ggplot(dat, aes(x = group, y = Ytilde))
  } else {
    p <- ggplot(dat, aes(y = Ytilde))
  }
    p <- p +
      geom_boxplot() +
      geom_hline(yintercept = 0, color = COLOR_ZERO_LINE) +
      stat_summary(fun.y=mean, color="red", geom="point")
  print(p)

}
```

## Adjusted phenotype

The adjusted phenotype values (`Ytilde`) are adjusted for fixed effect covariates,
random effects, and heterogeneous residual variances (if applicable); they
should be approximately distributed as N(0,1). The square of these values should
have a trend line near the y = 1 line, indicating homoskedasticity after
adjustment. Deviation from this trend could indicate that the variance model is
inadequate.


```{r adjusted-phenotype-plots}

p <- ggplot(dat, aes(x = fitted.values, y = Ytilde^2)) +
  geom_point(alpha = 0.1) +
  geom_hline(yintercept = 1, color = COLOR_ZERO_LINE) +
  geom_smooth(method = 'lm', color = 'red', fill = 'red') +
  geom_smooth(color = 'blue', linetype = 'dashed', fill = 'blue')
print(p)

p <- p +
  coord_cartesian(ylim = c(0, 5)) +
  ylab("Ytilde^2 zoomed in")
print(p)
```

The projected phenotype values (`PY`) are the phenotype values after adjusting
for the fixed effect covariates, random effects, and heterogenous residual
variances (if applicable); these are the values that are correlated with
genotype to test for association.

```{r, results = 'asis'}
if (!is_binary & !params$invnorm) {
  text <- paste(
    "It may be problematic for association testing if this distribution is far
    from Gaussian. This problem can be alleviated by rank normalizing `PY` and
    re-fitting the null model with the rank-normalized values as the outcome
    (i.e. a two stage procedure)."
  )
  cat(text)
}
```

```{r}
p <- ggplot(dat, aes(x = PY))
if (!is.null(group_var)) {
  p <- p +
    geom_density(aes(color = group)) +
    geom_density(size = 2)
} else {
  p <- p +
    geom_histogram()

}
print(p)
```

# Session information

```{r sessionInfo(), results = 'markup'}
print(sessionInfo())
```
